Extreme rainfall in Paraguay during the 2015-2016 austral summer: causes and subseasonal-to-seasonal predictive skill

### 2017-09-21 writeup/paper.tex
Re-vamp figures

Different types of analyses are used to diagnose the causes of the extreme rainfall events, and to bias-correct and verify the forecasts.
All codes used for data and analyis are available at \url{https://github.com/jdossgollin/PYFloods}.
Computation with multidimensional data was performed using the \texttt{xarray} \citep{hoyer2017xarray} and \texttt{numpy} \citep{vanderWalt:2011dp} packages of the python computing environment; computations on tabular data utilized the \texttt{pandas} \citep{McKinney:2010un} package; and plotting was performed with the \texttt{matplotlib} \citep{Hunter:2007ux} and \texttt{cartopy} \citep{Cartopy} packages.

@@

The $k$-means algorithm is guaranteed to converge to at least a local minimum; to select the best partition, 500 simulations were created using the implementation in Python's \texttt{scikit-learn} package \citep{Pedregosa:2012tv}.
The classifiability index of \citet{Michelangeli1995} was then used to select the optimal partition.
Calculation of the classifiability index for several values of $k$ (supplemental figure S2) suggests that states with $k=4,5,6,7,8$ are all reasonable; we chose the solution $k=6$ because the weather types identified are qualitatively similar to those determined over southeastern South America \citep{Munoz2015,Munoz2016} and have an intuitive physical meaning, discussed further in the following sections.

@@

\Cref{fig:lagged-rain} shows time-lagged anomalies up to and after days exceeding the 95th percentile of area-averaged daily rainfall in the lower Paraguay River basin (see \cref{fig:study-area}), and is consistent with the findings of \citet{Marengo2004,Salio:2007gd}.
At $t=-2$ and $t=-1$ a low-level northerry jet injects inflow of heat and moisture to the region, favored by a polar trough extending slightly to the east of Paraguay.
This is in agreement with previous studies which find a close correspondence between the exit regions of the low-level jet and the observed rainfall \citep{Saulo:2007km,Salio:2007gd,Marengo2004,Velasco1987} as the moisture influx, energy influx, and strong vertical wind shear favor instability and intense convection \citep{Marengo2004,Silva2009}.
Next, low-level convergence and upper-level divergence (not shown) associated with the eastward progression of the frontal system further favor rainfall locally.
Similar analysis for the 90th or 99th percentiles of daily rainfall (not shown) yield highly similar results, implying that the mechanism for the most extreme events is not fundamentally distinct from the mechanism for moderate-intensity events.

@@

We next use the weather typing algorithm to understand circulations and sequences of circulations associated with intense rainfall in the Lower Paraguay River Basin.

@@

Extreme rainfall over the region was accurately predicted by the IRI for the DJF 2015-2016 season since at least November 2015 (see \cref{fig:subs-prob-fcst}). Probabilities of 30\%-50\% for the 90th percentile, three to five times the expected climatological values, are visible over southern Paraguay and Brazil, and northern Uruguay and Argentina, mostly in agreement with observations (hashed region in \cref{fig:probfcst}).
Although this seasonal prediction was not skillful over eastern Paraguay, the forecast  showed a clear signal in the neighborhood, that could have been used for disaster-preparedness weeks in advance of the events.

@@

As seen in \Cref{fig:subs-prob-fcst}a, the raw (uncorrected) sub-seasonal forecast of the ECMF model for Dec 1-7 2015, indicated very high odds for occurrence of extreme rainfall but with important biases in the actual location and spatial pattern; for Paraguay, it confidently suggests occurrence of extremes to the south-southeast of the country, which was mostly not observed. Overall, the 20-year based skill of probabilistic forecasts for the first week of December --even when using multiple initializations-- is worse than using climatology for most of Paraguay (\Cref{fig:subs-prob-fcst}b), being best for the zone between that country and northern Uruguay, and for most of southern Brazil.

@@

Up to this point I think that we do a good job presenting what happened, and how the circulation anomalies led to rainfall.
I think the first piece of the discussion should focus on \emph{why} the observed circulation anomalies occurred.
\begin{itemize}
	\item I think this is the appropriate place to discuss ENSO and MJO teleconnections --the interaction of the two is very consistent with observed rainfall and circulation patterns
	\item Compared to other El Ni\~{n}o years, there were a lot of similarities, but also some very important differences including increased upper-level divergence over Paraguay, and
	\item \citet{Grimm2009} argues that land-surface interactions are important for the lifecycle of ENSO's impact on the region.
	Since the 2015-16 El Ni\~no was super strong, and 2014-15 was a weak EN year, it seems likely that these land-surface interactions and land-ocean temperature gradients were different amongst EN years historically; we don't know that this *is* the mechanism, but it would be interesting to explore further. MENTION THIS BUT I SUGGEST NOT TO FURTHER EXPLORE IT HERE. WE HAVE ENOUGH.
	\item By examining \cref{fig:klee}, we can see that some weather types tend be more likely to occur during specific parts of the 120-day season; this is consistent with \citet{Munoz2016} and with the notion that there is some structure to the nonlinear dynamical system, which could support the land-surface feedbacks that are proposed by i.e. \citep{Grimm:2009bq}.
\end{itemize}

@@

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../_figs/StudyArea.pdf}
	\caption{
		Map of the study area.
		(L): all of South America.
		The domains of the ``Lower Paraguay River Basin'' and the domain used for weather typing are indicated in black and red, respectively.
		(R): The Paraguay River.
		As for (L), the ``Lower Paraguay River Basin'' is indicated.
		Streamflows shown in \cref{fig:streamflow} were taken from the fours stations indicated in green.
		The Paraguay River and its tributaries, from the Natural Earth database (\texttt{www.naturalearthdata.com}), are also shown.
	}
  \label{fig:study-area}
\end{figure*}

@@

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../_figs/NDJF201516Anomaly.pdf}
	\caption{
    	Monthly composite anomalies observed during NDJF 2015-16.
      Top row shows anomalies of streamfunction $\psi$ at \SI{850}{\hecto\pascal}.
      Bottom row shows monthly rainfall anomalies, in units of \si{\milli\meter\per\day}.
	}
  \label{fig:anomalies}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../_figs/PSI_EOF_Loadings.pdf}
	\caption{
    	Leading four EOF loadings of daily NDJF \SI{850}{\hecto\pascal} streamfunction $\psi$ over chosen domain.
	}
  \label{fig:eof-loading}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in,height=9in,keepaspectratio=true]{../_figs/WTComposite.pdf}
	\caption{
		Composite anomalies associated with each weather type (column).
    Top row shows anomalies of streamfunction $\psi$ at \SI{850}{\hecto\pascal}.
    Bottom row shows precipitation anomalies in units of \si{\milli\meter\per\day}.
	}
	\label{fig:wt-composite}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../_figs/RainfallWeatherType}
	\caption{
		Time series of area-averaged rainfall in the Lower Paraguay River Basin (\cref{fig:study-area}) for each day of NDJF  2015-16 (lines).
		The weather types corresponding to each day are indicated in both color and adjacent label.
		Dashed lines indicate the climatological 50th, 90th, and 99th percentiles of NDJF area-averaged rain over the Lower Paragauy River basin.
	}
  \label{fig:rain-wt}
\end{figure*}

\begin{figure*}	\noindent\includegraphics[width=6.5in]{../_figs/LaggedRain.pdf}
	\caption{
		Circulation anomalies associated with intense rainfall (95th percentile exceedance of area-averaged rainfall in the Lower Paraguay River Basin shown in \cref{fig:study-area}).
		Lagged composites are shown, by column, for $t = $ \SIlist{-2;-1;0;1}{\day} relative to the date of intense rainfall.
		Top row shows anomalies of streamfunction $\psi$ at \SI{850}{\hecto\pascal}.
    Bottom row shows precipitation anomalies in units of \si{\milli\meter\per\day}.
	}
  \label{fig:lagged-rain}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../_figs/Chiclet.pdf}
	\caption{
    S2S model forecasts of area-averaged rainfall over the Lower Paraguay River Basin (\cref{fig:study-area}).
    Forecast time series from the ECMWF S2S model are shown for several lead times (\SIlist{5;14;19;30}{\day}).
		Each gray dot represents the forecast of a single ensemble member issued for the target date ($x$-axis) of area-averaged rainfall, in \si{\milli\meter\per\day} ($y$-axis).
		The blue line connects the ensemble means, and the black line shows the observed rainfall.
		Because the model is initialized twice-weekly, for a given lead time forecasts will only exist for two in seven possible target dates;  consequently the rainfall is smoothed with a 3-day running mean to facilitate visual comparison.
	}
  \label{fig:chiclet}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.4in]{../_figs/ForecastSkill.pdf}
	\caption{
		MOS-adjusted S2S model forecasts and skill scores for the methods indicated in \cref{tab:mos-methods}.
		Top row shows the intense rainfall forecast for 1-7 December 2015 as the odds ratio of \cref{eq:odds-ratio} over the target domain.
		A value greater than 1 indicates that the model forecast greater-than-average odds of rainfall exceeding the 90th percentile.
		The second shows the 2AFC skill score for each grid cell; a value greater than 50 indicates that the model outperforms climatology.
	}
  \label{fig:subs-prob-fcst}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../_figs/SeasonalForecast.pdf}
	\caption{
	Seasonal model forecast.
	Shading indicates the forecast probability of exceeding the 90th percentile of climatological rainfall during DJF 2015-16.
	As \cref{fig:subs-prob-fcst}, the variable shown is the odds ratio of \cref{eq:odds-ratio} over the target domain.
	A value greater than 1 indicates that the model forecast greater-than-average odds of rainfall exceeding the 90th percentile.
	}
  \label{fig:seas-prob-fcst}
\end{figure*}

>>>>

Different types of analyses are used to diagnose the causes of the extreme rainfall events, and to bias-correct and verify the forecasts.

All codes used for data and analyis are available at \url{https://github.com/jdossgollin/PYFloods}.
Computation with multidimensional data was performed using the \texttt{xarray} \citep{hoyer2017xarray} and \texttt{numpy} \citep{vanderWalt:2011dp} packages of the python computing environment; computations on tabular data utilized the \texttt{pandas} \citep{McKinney:2010un} package; and plotting was performed with the \texttt{matplotlib} \citep{Hunter:2007ux} and \texttt{cartopy} \citep{Cartopy} packages.

@@

The $k$-means algorithm is guaranteed to converge to at least a local minimum; to select the best partition, 250 simulations were created using the implementation in Python's \texttt{scikit-learn} package \citep{Pedregosa:2012tv}.
The classifiability index of \citet{Michelangeli1995} was then used to select the optimal partition.
Calculation of the classifiability index for several values of $k$ (supplemental figure S2) suggests that states with $k=4,5,6,7,8$ are all reasonable; we chose the solution $k=6$ because the weather types identified are qualitatively similar to those determined over southeastern South America \citep{Munoz2015,Munoz2016} and have an intuitive physical meaning, discussed further in the following sections.

@@

\Cref{fig:lagged-rain} shows time-lagged anomalies up to and after days exceeding the 95th percentile of area-averaged daily rainfall in the lower Paraguay River basin (see \cref{fig:study-area}), and is consistent with previous findings \citep{Marengo2004,Salio:2007gd}.
At $t=-2$ a mid-latitude baroclinic system approaches the South American continent, intensifying and moving to the East at $t=-1,0,1$.
Along the cold front associated with this system, a low-level northerly jet advects of heat and moisture to the region, favored by a polar trough extending slightly to the east of Paraguay.
This is in agreement with previous studies which find a close correspondence between the exit regions of the low-level jet and the observed rainfall \citep{Saulo:2007km,Salio:2007gd,Marengo2004,Velasco1987} as the moisture influx, energy influx, and strong vertical wind shear favor instability and intense convection \citep{Marengo2004,Silva2009}.
Next, low-level convergence and upper-level divergence (not shown) associated with the eastward progression of the frontal system further favor rainfall over the Lower Paraguay River Basin.
As the system progresses, the rainfall anomaly approximately follows the anomaly of the low-level wind.

Similar analysis for the 90th or 99th percentiles of daily rainfall (not shown) yield similar results, implying that the synoptic mechanism for the most extreme events is not fundamentally distinct from the mechanism for moderate-intensity events.

@@

We next use the weather typing algorithm to understand particular circulations and sequences of circulations associated with intense rainfall in the Lower Paraguay River Basin.

@@

Extreme rainfall over the region was accurately predicted by the IRI for the DJF 2015-2016 season since at least November 2015 (see \cref{fig:seas-prob-fcst}). Probabilities of 30\%-50\% for the 90th percentile, three to five times the expected climatological values, are visible over southern Paraguay and Brazil, and northern Uruguay and Argentina, mostly in agreement with observations (hashed region in \cref{fig:seas-prob-fcst}).
Although this seasonal prediction was not skillful over eastern Paraguay, the forecast  showed a clear signal in the neighborhood, that could have been used for disaster-preparedness weeks in advance of the events.

@@

As seen in \Cref{fig:subs-prob-fcst}a, the raw (uncorrected) sub-seasonal forecast of the ECMF model for Dec 1-7 2015, indicated very high odds for occurrence of extreme rainfall but with important biases in the actual location and spatial pattern; for Paraguay, it confidently suggests occurrence of extremes to the south-southeast of the country, which was mostly not observed. Overall, the 20-year based skill of probabilistic forecasts for the first week of December --even when using multiple initializations-- is worse than using climatology for most of Paraguay (\cref{fig:subs-prob-fcst}b), being best for the zone between that country and northern Uruguay, and for most of southern Brazil.

@@

Up to this point I think that we do a good job presenting what happened, and how the circulation anomalies led to rainfall.
I think the first piece of the discussion should focus on \emph{why} the observed circulation anomalies occurred.
\begin{itemize}
	\item I think this is the appropriate place to discuss ENSO and MJO teleconnections --the interaction of the two is very consistent with observed rainfall and circulation patterns
	\item Compared to other El Ni\~{n}o years, there were a lot of similarities, but also some very important differences including increased upper-level divergence over Paraguay, and
	\item \citet{Grimm2009} argues that land-surface interactions are important for the lifecycle of ENSO's impact on the region.
	Since the 2015-16 El Ni\~no was super strong, and 2014-15 was a weak EN year, it seems likely that these land-surface interactions and land-ocean temperature gradients were different amongst EN years historically; we don't know that this *is* the mechanism, but it would be interesting to explore further. MENTION THIS BUT I SUGGEST NOT TO FURTHER EXPLORE IT HERE. WE HAVE ENOUGH.
	\item By examining \cref{fig:rain-wt}, we can see that some weather types tend be more likely to occur during specific parts of the 120-day season; this is consistent with \citet{Munoz2016} and with the notion that there is some structure to the nonlinear dynamical system, which could support the land-surface feedbacks that are proposed by i.e. \citep{Grimm:2009bq}.
\end{itemize}

@@

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../figs/StudyArea.pdf}
	\caption{
		Map of the study area.
		(L): all of South America.
		The domains of the ``Lower Paraguay River Basin'' and the domain used for weather typing are indicated in black and red, respectively.
		(R): The Paraguay River.
		As for (L), the ``Lower Paraguay River Basin'' is indicated.
		Streamflows shown in \cref{fig:streamflow} were taken from the fours stations indicated in green.
		The Paraguay River and its tributaries, from the Natural Earth database (\texttt{www.naturalearthdata.com}), are also shown.
	}
  \label{fig:study-area}
\end{figure*}

@@

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../figs/NDJF201516Anomaly.pdf}
	\caption{
    	Monthly composite anomalies observed during NDJF 2015-16.
      Top row shows anomalies of streamfunction $\psi$ at \SI{850}{\hecto\pascal}.
      Bottom row shows monthly rainfall anomalies, in units of \si{\milli\meter\per\day}.
	}
  \label{fig:anomalies}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../figs/PSI_EOF_Loadings.pdf}
	\caption{
    	Leading four EOF loadings of daily NDJF \SI{850}{\hecto\pascal} streamfunction $\psi$ over chosen domain.
	}
  \label{fig:eof-loading}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in,height=9in,keepaspectratio=true]{../figs/WTComposite.pdf}
	\caption{
		Composite anomalies associated with each weather type (column).
    Top row shows anomalies of streamfunction $\psi$ at \SI{850}{\hecto\pascal}.
    Bottom row shows precipitation anomalies in units of \si{\milli\meter\per\day}.
	}
	\label{fig:wt-composite}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../figs/RainfallWeatherType}
	\caption{
		Time series of area-averaged rainfall in the Lower Paraguay River Basin (\cref{fig:study-area}) for each day of NDJF  2015-16 (lines).
		The weather types corresponding to each day are indicated in both color and adjacent label.
		Dashed lines indicate the climatological 50th, 90th, and 99th percentiles of NDJF area-averaged rain over the Lower Paragauy River basin.
	}
  \label{fig:rain-wt}
\end{figure*}

\begin{figure*}	\noindent\includegraphics[width=6.5in]{../figs/LaggedRain.pdf}
	\caption{
		Circulation anomalies associated with intense rainfall (95th percentile exceedance of area-averaged rainfall in the Lower Paraguay River Basin shown in \cref{fig:study-area}).
		Lagged composites are shown, by column, for $t = $ \SIlist{-2;-1;0;1}{\day} relative to the date of intense rainfall.
		Top row shows anomalies of streamfunction $\psi$ at \SI{850}{\hecto\pascal}.
    Bottom row shows precipitation anomalies in units of \si{\milli\meter\per\day}.
	}
  \label{fig:lagged-rain}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../figs/Chiclet.pdf}
	\caption{
    S2S model forecasts of area-averaged rainfall over the Lower Paraguay River Basin (\cref{fig:study-area}).
    Forecast time series from the ECMWF S2S model are shown for several lead times (\SIlist{5;14;19;30}{\day}).
		Each gray dot represents the forecast of a single ensemble member issued for the target date ($x$-axis) of area-averaged rainfall, in \si{\milli\meter\per\day} ($y$-axis).
		The blue line connects the ensemble means, and the black line shows the observed rainfall.
		Because the model is initialized twice-weekly, for a given lead time forecasts will only exist for two in seven possible target dates;  consequently the rainfall is smoothed with a 3-day running mean to facilitate visual comparison.
	}
  \label{fig:chiclet}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.4in]{../figs/ForecastSkill.pdf}
	\caption{
		MOS-adjusted S2S model forecasts and skill scores for the methods indicated in \cref{tab:mos-methods}.
		Top row shows the intense rainfall forecast for 1-7 December 2015 as the odds ratio of \cref{eq:odds-ratio} over the target domain.
		A value greater than 1 indicates that the model forecast greater-than-average odds of rainfall exceeding the 90th percentile.
		The second shows the 2AFC skill score for each grid cell; a value greater than 50 indicates that the model outperforms climatology.
	}
  \label{fig:subs-prob-fcst}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../figs/SeasonalForecast.pdf}
	\caption{
	Seasonal model forecast.
	Shading indicates the forecast probability of exceeding the 90th percentile of climatological rainfall during DJF 2015-16.
	As \cref{fig:subs-prob-fcst}, the variable shown is the odds ratio of \cref{eq:odds-ratio} over the target domain.
	A value greater than 1 indicates that the model forecast greater-than-average odds of rainfall exceeding the 90th percentile.
	}
  \label{fig:seas-prob-fcst}
\end{figure*}
